### 2025-06-21 docs/thesis/main.tex
Improve text
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\pagenumbering{roman}
\input{frontmatter/frontpage.tex}
\pagecolor{white}
\newgeometry{top=2.81cm, bottom=2.75cm, outer=2.5cm, inner=3.5cm}
\pagestyle{empty}
\cleardoublepage 
\input{frontmatter/copyright.tex}
\clearpage 
\pagestyle{main}
\input{frontmatter/approval.tex}
\clearpage 
\input{frontmatter/abstract.tex}
\clearpage 
\input{frontmatter/acknowledgements.tex}
\cleardoublepage 
\tableofcontents
\cleardoublepage

@@

We evaluate \sys across a diverse set of transformation tasks~(\cref{sec:evaluation}), including:
	security hardening, by stripping latent malicious behavior while preserving intended functionality;
	cross-language translation, by regenerating semantically equivalent programs in new languages;
	idiomatic rewriting, by producing safer and more portable variants of legacy code;
	modular decomposition, by restructuring monolithic components into composable parts.
\sys is able to effectively regenerate programs across these tasks, producing high-quality outputs that satisfy user-specified constraints while also discovering transformation strategies that were not immediately obvious to the authors.

@@

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
graph(P_, I, O) :- graph(P, I, O).
:- not pure(P_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\texttt{input(P, I), graph(P, I, O), signature(P, S), pure(P)}\}$, where
\texttt{input(P, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.
The, \texttt{pure(P)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.

@@

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
equal(graph(P), graph(P_)).
language(P_, haskell).
\end{minted}
\end{wrapfigure}
The query requests preservation of I/O behavior and translation to Haskell. The
logic engine deduces that extracting I/O traces and the function signature is
sufficient; the knowledge base automatically associates the target language
\texttt{Haskell} with purity.
\sys transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\texttt{([1,[2,3]], (x)=>x+1)}\to\texttt{[2,3,4]}\rangle$ 
becomes $\langle(\texttt{[1, [2,3]], Js("(x)=>x+1"))}\to\texttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated 20 I/O pairs, \sys produces the following
Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[6]{r}{.7\columnwidth}
  \vspace{-10pt}
  \begin{minted}{prolog}
db('music.db').
trace_sql(p_, I, T) :- trace_sql(p, I, T),
                       graph(p, I, _).
func(f1). func(f2).
graph(call(f2, call(f1), I), I, O) :- graph(p, I, O).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database) driven by an input.
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for 24 generated inputs.
Then, after 20.4 seconds and two attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

@@

We evaluate \sys along three axes:

@@

We evaluate \sys on six benchmark suites, summarized in \cref{tab:benchmarks}.
Each task consists of an input program, a \sys query expressing desired
properties, and a test suite or functional oracle.
Each benchmark is executed three times. A regeneration is considered overal succesful 
if it is correct, satisfies the query, and passes all tests all three times.
The Rosetta Code~\cite{rosettacode} benchmarks include four programming tasks written in four different
languages each (Python, JavaScript, C, and Haskell). 
These include classic programs that do not require any language-specific features to implement.
These are: the classic \emph{Hello, World!} program as found in most programming tutorials,
a program that applies the ROT-13 cipher to a string,
a program that computes the median of a list of numbers,
and a program that reads a file and prints its contents to the console line-by-line.
The npm benchmark includes ten popular (100.000+ weekly downloads) npm packages that implement utility
functionality such as string manipulation, data structure transformations, and math.
Similalry, the PyPi utilities benchmark includes three functions sourced from PyPi that implement similar functionality.
The SSCA dataset includes three sneaky supply-chain attacks that obfuscate malicious behavior in otherwise innocuous-looking code.
The attack activates under specific environment conditions, such as the presence of a specific environment variable or specific inputs
getting passed to the program. The attacks involve private key exfiltration~\cite{ohm2020backstabber}, Bitcoin private key theft~\cite{ev:eurosec:2022}, and an SSH backdoor installation~\cite{copeland2019frightening}.
and occur while the program still performs its advertised functionality.
The IOCCC benchmark includes two programs from the International Obfuscated C Code Contest (IOCCC)~\cite{ioccc}.
These programs are intentionally obfuscated, using advanced techniques such as bit manipulation, undefined behavior,
and exploitation of the C preprocessor and the lax parsing rules of the C language.
These programs perform mathematical computations like mersense primality testing and adding two numbers.
Finally, the microbenchmarks include three programs that implement the regeneration tasks discussed in \cref{sec:example}.

@@

We evaluate correctness using developer-made test suites 
shipped with the original software components as well as by manually inspecting
the regenerated source code and report on correctness and the other query target metrics.
When no test suite is available, we only do manual inspection.

\begin{table}[h]
  \centering
  \caption{Correctness: Percentage of tasks passing all tests (when present) and manual inspection.}
  \begin{tabular}{lc}
    \toprule
    Benchmark & \sys Pass Rate \\ 
    \midrule
    Rosetta Code & 16/16  \\
    Short Utilities & 13/10 \\
    SSCA & 3/3  \\
    IOCCC & 1/2 \\
    Microbenchmarks & 2.5/3  \\
    \bottomrule
  \end{tabular}
\end{table}

\sys succeeds in all of the Rosetta Code benchmarks, code is correct, we had to manually verify it as no test cases are available.
\sys regenerates correctly all npm and PyPi utilities, it does not introduce and sideffectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \texttt{left-pad} library, precomputes padding for strings below a certain length
and returns it directly. In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge.
For example, in the \texttt{is} family of libraries,
\sys-generated code frequently fails to account for all edge cases,
and often misses JavaScript-specific patterns,
such as checking an object’s string representation to decide if an object is circular, as in the \texttt{is-circular} library.
\sys has weak performance on libraries that involve complex regexes or perform non-trivial math.
For the SSCA benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses some domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} plugin) to generate a function that produces a node \ttt{Stream} object.
For the IOCCC benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the semantics of the original one, as it operates over infinitely nested lists, rather than only one level deep ones.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic computation of the inverse square root ($1/\sqrt{x}$) rather than the original bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

\promptbox{
  \emph{Task description} \eg {Give back the same code having removed any vulnuerabilities}\\

@@

\sys consistently outperforms a \gptmodel baseline,
highlighting the inherent weaknesses of a purely prompt-based approach.

@@

\sys consistently satisfies more of the declared properties while avoiding
unintended behaviors. For instance, in the \texttt{flatmap-stream} benchmark,
\gptmodel often inclduded malicious side effects when the code appeared innocuous,
whereas \sys generated pure rewrites by treating the source as
untrusted and grounding regeneration in I/O behavior alone.
Across all benchmarks, the planning step completes in under 50ms on average.
Similalry, in the \ttt{Q\_rsqrt} benchmark, \gptmodel often produces bit-level
manipulations that are not idiomatic in Haskell, while \sys generates a simple
and idiomatic Haskell function that computes the inverse square root.
This is guided by the fact the \sys extracts and provides to its backend model only the function signature 
and length of the program, which it minimizes, without including the original source code,
which as shown in the experimental results makes the model overfit its regeneration to it.
\sys's I/O pair extraction also provides confidence that 
the regenerated program has the same client-visible behavior as the original one, while 
prompting a model like \gptmodel would always require manual inspection or the presence 
of developer tests to verify the correctness of the regeneration.

These results support the claim that \sys's planning logic---driven by explicit
property extraction and knowledge-base reasoning---yields safer and more
controllable transformations than prompt-only LLM-based systems.

@@

\heading{Future directions}
Extensions that could improve \sys's capabilities include richer property languages (\eg temporal or dataflow properties~\cite{azzopardi2023ltl,handa2021orderawaredataflowmodelparallel}), tighter integration with formal verification tools, and cross-component reasoning for coordinated rewrites.
Furthermore, performance-oriented transformations could be potent additions, allowing \sys to transform the program across dimensions like runtime, parallelizability, or memory usage.
Additionally, integrating counterfactual generation techniques~\cite{Cabalar_2020} could help explain why certain regeneration goals are infeasible or conflicting.

@@

\sys opens the door to a new class of synthesis tools that combine symbolic
reasoning and neural generation under a unified, extensible framework.
As program synthesis systems grow increasingly reliant on LLMs, we believe the
future lies in harnessing their generative power within a framework that retains
correctness, explainability, and developer control as first-class citizens.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\pagenumbering{roman}
\input{frontmatter/frontpage.tex}
\pagecolor{white}
\newgeometry{top=2.81cm, bottom=2.75cm, outer=2.5cm, inner=3.5cm}
\pagestyle{empty}
\cleardoublepage 
\input{frontmatter/copyright.tex}
\clearpage 
\pagestyle{main}
\input{frontmatter/approval.tex}
\clearpage 
\input{frontmatter/abstract.tex}
\clearpage 
% \input{frontmatter/acknowledgements.tex}
% \cleardoublepage 
\tableofcontents
\cleardoublepage

@@

\sys is evaluated across a diverse set of transformation tasks~(\cref{sec:evaluation}), including:
	security hardening, by stripping latent malicious behavior while preserving intended functionality;
	cross-language translation, by regenerating semantically equivalent programs in new languages;
	idiomatic rewriting, by producing safer and more portable variants of legacy code;
	modular decomposition, by restructuring monolithic components into composable parts.
\sys is able to effectively regenerate programs across these tasks, producing high-quality outputs that satisfy user-specified constraints while also discovering transformation strategies that were not immediately obvious to the authors.

@@

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\texttt{input(P, I), graph(P, I, O), signature(P, S), pure(P)}\}$, where
\texttt{input(P, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.
The, \texttt{pure(P)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.

@@

\begin{wrapfigure}[3]{r}{.5\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query requests preservation of I/O behavior and translation to Haskell. The
logic engine deduces that extracting I/O traces and the function signature is
sufficient; the knowledge base automatically associates the target language
\texttt{Haskell} with purity.
\sys transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\texttt{([1,[2,3]], (x)=>x+1)}\to\texttt{[2,3,4]}\rangle$ 
becomes $\langle(\texttt{[1, [2,3]], Js("(x)=>x+1"))}\to\texttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated 20 I/O pairs, \sys produces the following
Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[5]{r}{.85\columnwidth}
  \begin{minted}{prolog}
db('music.db'). func(f1). func(f2).
trace_sql(p_, I, T) :- trace_sql(p, I, T), graph(p, I, _).
graph(call(f2, call(f1), I), I, O) :- graph(p, I, O).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database) driven by an input.
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for 24 generated inputs.
Then, after 20.4 seconds and two attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\heading{Answer Set Programming}
Answer Set Programming (ASP) \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo} \cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

The syntax, similar to that of \textit{Prolog} and \textit{Datalog}, with code and data being represented
through logical terms. Collections of logical terms, as seen in
\cref{lst:terms}, can represent the game world's state. Rules, expressed using
the \texttt{:-} operator, enable complex reasoning by defining relationships
between atoms. Choice rules, exemplified in \cref{lst:choice}, allow the ASP
solver to make selections among atoms. Integrity constraints, as shown in
\cref{lst:constraint}, can restrict invalid answer sets. Optimization
directives, such as \texttt{\#minimize} and \texttt{\#maximize} in
\cref{lst:optimization}, guide the solver to output optimal answer sets.

\begin{lstlisting}[caption={A set of facts over a program.}, label=lst:terms]{Name}
language(p, javascript).

\end{lstlisting}

\begin{lstlisting}[caption={Logic rules describing relationships between the original and regenerated program.}]{Name}
graph(p_) :- graph(p).
len(p_, N) :- len(p, N_), N = N_ + 1.
\end{lstlisting}

\begin{lstlisting}[caption={A choice rule.}, label=lst:choice]{Name}
{absent(F) : function(F, P)} :- program(P).
\end{lstlisting}

\begin{lstlisting}[caption={An integrity constraint.}, label=lst:constraint]{Name}
:- language(p, X), language(p_, Y), X != Y.
\end{lstlisting}

\begin{lstlisting}[caption={An optimization directive.}, label=lst:optimization]{Name}
#minimize{C : cost(E,C)}.
\end{lstlisting}

% To compute answer sets, ASP programs are input into ASP solvers.
% These solvers provide efficient mechanisms to generate the set of valid answers to the given problem.
% An ASP solver can be thought of as a black box, with them being interchangeable as 
% long as the input language semantics remain the same.

@@

The \sys evaluation is along three axes:

@@

The \sys evaluation comprises of six benchmark suites, summarized in \cref{tab:benchmarks}.
Each task consists of an input program, a \sys query expressing desired
properties, and a test suite or functional oracle.
Each benchmark is executed three times. A regeneration is considered overal succesful 
if it is correct, satisfies the query, and passes all tests all three times.
The Rosetta Code~\cite{rosettacode} benchmarks include four programming tasks written in four different
languages each (Python, JavaScript, C, and Haskell). 
These include classic programs that do not require any language-specific features to implement.
These are: the classic \emph{Hello, World!} program as found in most programming tutorials,
a program that applies the ROT-13 cipher to a string,
a program that computes the median of a list of numbers,
and a program that reads a file and prints its contents to the console line-by-line.
The npm benchmark includes ten popular (100.000+ weekly downloads) npm packages that implement utility
functionality such as string manipulation, data structure transformations, and math.
Similalry, the PyPi utilities benchmark includes three functions sourced from PyPi that implement similar functionality.
The SSCA dataset includes three sneaky supply-chain attacks that obfuscate malicious behavior in otherwise innocuous-looking code.
The attack activates under specific environment conditions, such as the presence of a specific environment variable or specific inputs
getting passed to the program. The attacks involve private key exfiltration~\cite{ohm2020backstabber}, Bitcoin private key theft~\cite{ev:eurosec:2022}, and an SSH backdoor installation~\cite{copeland2019frightening}.
and occur while the program still performs its advertised functionality.
The IOCCC benchmark includes two programs from the International Obfuscated C Code Contest (IOCCC)~\cite{ioccc}.
These programs are intentionally obfuscated, using advanced techniques such as bit manipulation, undefined behavior,
and exploitation of the C preprocessor and the lax parsing rules of the C language.
These programs perform mathematical computations like mersense primality testing and adding two numbers.
Finally, the microbenchmarks include three programs that implement the regeneration tasks discussed in \cref{sec:example}.

@@

Correctness is evaluated using developer-made test suites 
shipped with the original software components as well as by manually inspecting
the regenerated source code and report on correctness and the other query target metrics.
When no test suite is available, only manual inspection is performed.

\begin{table}[h]
  \centering
  \caption{Correctness: Number of tasks passing all tests (when present) and manual inspection.}
  \begin{tabular}{lc}
    \toprule
    Benchmark & Correctness \\ 
    \midrule
    Rosetta Code & 16/16  \\
    Short Utilities & 13/10 \\
    SSCA & 3/3  \\
    IOCCC & 1/2 \\
    Microbenchmarks & 2.5/3  \\
    \bottomrule
  \end{tabular}
\end{table}

\sys successfully regenerates all programs in the Rosetta Code benchmark suite,
having verified manually that the output is functionally equivalent
to the original, using both the original and translated reference
implementations as guides.
\sys regenerates correctly all npm and PyPi utilities, it does not introduce and sideffectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \texttt{left-pad} library, precomputes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge.
For example, in the \texttt{is} family of libraries,
\sys-generated code frequently fails to account for all edge cases.
% \sys has weak performance on libraries that involve complex regexes or perform non-trivial math.
For the SSCA benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses some domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} plugin) to generate a function that produces a node \ttt{Stream} object.
For the IOCCC benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the semantics of the original one, as it operates over infinitely nested lists, rather than only one level deep ones.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic computation of the inverse square root ($1/\sqrt{x}$) rather than the original bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

\promptbox{
  \emph{Task description}\\

@@

\sys consistently outperforms a \gptmodel baseline,
overcoming the inherent weaknesses of a purely prompt-based approach.

@@

\sys consistently satisfies more of the declared properties while avoiding
unintended behaviors.
Following is a qualitative analysis of the results.
In the \texttt{flatmap-stream} benchmark,
\gptmodel often inclduded malicious side effects when the code appeared innocuous,
whereas \sys generated pure rewrites by treating the source as
untrusted and grounding regeneration in I/O behavior alone.
Across all benchmarks, the planning step completes in under 50ms on average.
Similalry, in the \ttt{Q\_rsqrt} benchmark, \gptmodel often produces bit-level
manipulations that are not idiomatic in Haskell, while \sys generates a simple
and idiomatic Haskell function that computes the inverse square root.
This is guided by the fact the \sys extracts and provides to its backend model only the function signature 
and length of the program, which it minimizes, without including the original source code,
which as shown in the experimental results makes the model overfit its regeneration to it.
\sys's I/O pair extraction also provides confidence that 
the regenerated program has the same client-visible behavior as the original one, while 
prompting a model like \gptmodel would always require manual inspection or the presence 
of developer tests to verify the correctness of the regeneration.

% These results support the claim that \sys's planning logic---driven by explicit
% property extraction and knowledge-base reasoning---yields safer and more
% controllable transformations than prompt-only LLM-based systems.

@@

@@

\sys opens the door to a new class of synthesis tools that combine symbolic
reasoning and neural generation under a unified, extensible framework.
As program synthesis systems grow increasingly reliant on LLMs, I believe the
future lies in harnessing their generative power within a framework that retains
correctness, explainability, and developer control as first-class citizens.

Here, I outline some possible future directions for \sys and its successors:

\heading{Expanding the transformation space}
Future extensions to \sys may incorporate richer classes of properties, such as
temporal logic and dataflow
constraints~\cite{azzopardi2023ltl,handa2021orderawaredataflowmodelparallel}.
In addition, performance-guided transformations offer a compelling avenue: \sys
could optimize for runtime, memory usage, or even parallelizability.
Inverse
transformations, such as serializing parallel programs, could serve as
a debugging or oracle mechanism in complex system settings.

\heading{Formal foundations and alternative implementations}
The transformation semantics introduced by \sys are grounded in logic
programming and inherit properties like compositionality.
A more formal treatment of the \sys
transformation language could enable rigorous reasoning about transformation
correctness and equivalence.
Furthermore, establishing an abstract interface
over transformations would allow alternative \sys backends to be developed.
Integrating techniques from
counterfactual reasoning~\cite{Cabalar_2020} could further enhance \sys by
explaining synthesis failures and clarifying conflicting property constraints.

\heading{Regeneration as a system-building primitive}
Beyond isolated transformations, \sys can serve as a core building block for
higher-level systems.
For instance, a reverse-engineering pipeline for
proprietary or legacy systems could use \sys to regenerate and recompose
well-behaved modules such as networking layer or the physics engine into new,
repurposed applications.
The vision here is \sys as a foundation for automated system understanding, adaptation, and reimagination.

\heading{Challenges in cross-language regeneration}
Cross-language regeneration introduces challenges.
Some analyses and
properties may only be applicable in a subset of languages, or lack direct
equivalents across language boundaries.
\sys could address this in two ways: by
relying on least-common denominator of language semantics, or by
embracing language-specific features to drive more effective regeneration.
In
fact, cross-compilation could be leveraged as an asset: a plugin pipeline might
convert JavaScript to Haskell, extract formal properties via Haskell’s type
system, and then regenerate a more robust JavaScript variant informed by that
analysis---effectively using \sys recursively.

\heading{Ethical and legal considerations}
While this work has applied \sys exclusively to open-source software, its
capabilities raise ethical and legal questions. Regenerating
proprietary software, deobfuscating binaries, or reimplementing licensed
systems into equivalent ones with more permissive licensing 
could challenge existing norms and
regulations. 
It is important to consider
safeguards and responsible disclosure practices towards ethical program regeneration.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
